# Causal Mediation Analysis

_Nima Hejazi_

Based on the [`tmle3mediate` `R`
package](https://github.com/tlverse/tmle3mediate).

## Learning Objectives

1. Understand that you can estimate the natural direct and indirect effects for a binary treatment using the `tmle3mediate` `R` package.
2. Know where to learn more about the `tmle3mediate` package.

## Causal Mediation Analysis

We only have time to briefly introduce these parameters. You can find more information in the relevant [handbook chapter](http://tlverse.org/tlverse-handbook/causal-mediation-analysis.html)

In the presence of post-treatment intermediate variables affected by
exposure (that is, _mediators_), path-specific effects allow for such complex,
mechanistic relationships to be teased apart. These causal effects are of such
wide interest that their definition and identification has been the object of
study in statistics for nearly a century -- indeed, the earliest examples of
modern causal mediation analysis can be traced back to work on path analysis
[@wright1934method]. 

In recent decades, renewed interest has resulted in the
formulation of novel direct and indirect effects within both the potential
outcomes and nonparametric structural equation modeling frameworks
[@robins1986new; @pearl1995causal; @pearl2009causality; @spirtes2000causation;
@dawid2000causal]. Generally, the indirect effect (IE) is the portion of the
total effect found to work _through_ mediating variables, while the direct
effect (DE) encompasses _all other components_ of the total effect, including
both the effect of the treatment directly on the outcome _and_ its effect
through all paths not explicitly involving the mediators. The mechanistic
knowledge conveyed by the direct and indirect effects can be used to improve
understanding of both _why_ and _how_ treatments may be efficacious.

## Data Structure and Notation

Let us return to our familiar sample of $n$ units $O_1, \ldots, O_n$, where we
now consider a slightly more complex data structure $O = (W, A, Z, Y)$ for any
given observational unit. As before, $W$ represents a vector of observed
covariates, $A$ a binary or continuous treatment, and $Y$ a binary or continuous
outcome; the new post-treatment variable $Z$ represents a (possibly
multivariate) set of mediators. Avoiding assumptions unsupported by background
scientific knowledge, we assume only that $O \sim P_0 \in \M$, where $\M$ is the
nonparametric statistical model that places no assumptions on the form of the
data-generating distribution $P_0$.

As in preceding chapters, a structural causal model (SCM) [@pearl2009causality]
helps to formalize the definition of our counterfactual variables:
\begin{align}
  W &= f_W(U_W) \\ \nonumber
  A &= f_A(W, U_A) \\ \nonumber
  Z &= f_Z(W, A, U_Z) \\ \nonumber
  Y &= f_Y(W, A, Z, U_Y).
  (\#eq:npsem-mediate)
\end{align}
This set of equations
constitutes a mechanistic model generating the observed data $O$; furthermore,
the SCM encodes several fundamental assumptions. Firstly, there is an implicit
temporal ordering: $W$ occurs first, depending only on exogenous factors $U_W$;
$A$ happens next, based on both $W$ and exogenous factors $U_A$; then come the
mediators $Z$, which depend on $A$, $W$, and another set of exogenous factors
$U_Z$; and finally appears the outcome $Y$. We assume neither access to the set
of exogenous factors $\{U_W, U_A, U_Z, U_Y\}$ nor knowledge of the forms of the
deterministic generating functions $\{f_W, f_A, f_Z, f_Y\}$. In practice, any
available knowledge about the data-generating experiment should be incorporated
into this model -- for example, if the data from a randomized controlled trial
(RCT), the form of $f_A$ may be known. The SCM corresponds to the following DAG:

```{r mediation-DAG, echo = FALSE, out.width = "60%"}
library(dagitty)
library(ggdag)

# make DAG by specifying dependence structure
dag <- dagitty(
  "dag {
    W -> A
    W -> Z
    W -> Y
    A -> Z
    A -> Y
    Z -> Y
    W -> A -> Y
    W -> A -> Z -> Y
  }"
)
exposures(dag) <- c("A")
outcomes(dag) <- c("Y")
tidy_dag <- tidy_dagitty(dag)

# visualize DAG
ggdag(tidy_dag) +
  theme_dag()
```

By factorizing the likelihood of the data $O$, we can express $p_0$, the
density of $O$ with respect to the product measure, when evaluated on a
particular observation $o$, in terms of several orthogonal components:
\begin{align}
  p_0(o) = &q_{0,Y}(y \mid Z = z, A = a, W = w) \\ \nonumber
    &q_{0,Z}(z \mid A = a, W = w) \\ \nonumber
    &g_{0,A}(a \mid W = w) \\ \nonumber
    &q_{0,W}(w).\\ \nonumber
  (\#eq:likelihood-factorization-mediate)
\end{align}
In Equation \@ref(eq:likelihood-factorization-mediate), $q_{0, Y}$ is the
conditional density of $Y$ given $\{Z, A, W\}$, $q_{0, Z}$ is the conditional
density of $Z$ given $\{A, W\}$, $g_{0, A}$ is the conditional density of $A$
given $W$, and $q_{0, W}$ is the marginal density of $W$. For convenience and
consistency of notation, we will define $\overline{Q}_Y(Z, A, W) := \E[Y \mid Z,
A, W]$ and $g(A \mid W) := \P(A \mid W)$ (i.e., the propensity score).

## Defining the Natural Direct and Indirect Effects

### Decomposing the Average Treatment Effect

The natural direct and indirect effects arise from a decomposition of the ATE:
\begin{align*}
  \E[Y(1) - Y(0)] =
    &\underbrace{\E[Y(1, Z(0)) - Y(0, Z(0))]}_{\text{NDE}} \\ &+
    \underbrace{\E[Y(1, Z(1)) - Y(1, Z(0))]}_{\text{NIE}}.
\end{align*}
In particular, the natural indirect effect (NIE) measures the effect of the
treatment $A \in \{0, 1\}$ on the outcome $Y$ through the mediators $Z$, while
the natural direct effect (NDE) measures the effect of the treatment on the
outcome _through all other pathways_.

Identifiability results and necessary assumptions for these parameters are thoroughly covered in the handbook.

### Estimating the Natural Direct Effect

The NDE is defined as
\begin{align*}
  \psi_{\text{NDE}} =&~\E[Y(1, Z(0)) - Y(0, Z(0))] \\
  =& \sum_w \sum_z
  [\underbrace{\E(Y \mid A = 1, z, w)}_{\overline{Q}_Y(A = 1, z, w)} -
  \underbrace{\E(Y \mid A = 0, z, w)}_{\overline{Q}_Y(A = 0, z, w)}] \\
  &\times \underbrace{p(z \mid A = 0, w)}_{q_Z(Z \mid 0, w))}
  \underbrace{p(w)}_{q_W},
\end{align*}
where the likelihood factors arise from a factorization of the joint
likelihood:
\begin{equation*}
  p(w, a, z, y) = \underbrace{p(y \mid w, a, z)}_{q_Y(A, W, Z)}
  \underbrace{p(z \mid w, a)}_{q_Z(Z \mid A, W)}
  \underbrace{p(a \mid w)}_{g(A \mid W)}
  \underbrace{p(w)}_{q_W}.
\end{equation*}

The process of estimating the NDE begins by constructing $\overline{Q}_{Y, n}$,
an estimate of the conditional mean of the outcome, given $Z$, $A$, and $W$.
With an estimate of this conditional mean in hand, predictions of the
quantities $\overline{Q}_Y(Z, 1, W)$ (setting $A = 1$) and,
likewise, $\overline{Q}_Y(Z, 0, W)$ (setting $A = 0$) are readily obtained. We
denote the difference of these conditional means $\overline{Q}_{\text{diff}} =
\overline{Q}_Y(Z, 1, W) - \overline{Q}_Y(Z, 0, W)$, which is itself only a
functional parameter of the data distribution. $\overline{Q}_{\text{diff}}$
captures differences in the conditional mean of $Y$ across contrasts of $A$.

A procedure for constructing a targeted maximum likelihood (TML) estimator of
the NDE treats $\overline{Q}_{\text{diff}}$ itself as a nuisance parameter,
regressing its estimate $\overline{Q}_{\text{diff}, n}$ on baseline covariates
$W$, among observations in the control condition only (i.e., those for whom
$A = 0$ is observed); the goal of this step is to remove part of the marginal
impact of $Z$ on $\overline{Q}_{\text{diff}}$, since the covariates $W$ precede
the mediators $Z$ in time. Regressing this difference on $W$ among the controls
recovers the expected $\overline{Q}_{\text{diff}}$, under the setting in which
all individuals are treated as falling in the control condition $A = 0$. Any
residual additive effect of $Z$ on $\overline{Q}_{\text{diff}}$ is removed
during the TML estimation step using the auxiliary (or "clever") covariate,
which accounts for the mediators $Z$. This auxiliary covariate takes the form

\begin{equation*}
  C_Y(q_Z, g)(O) = \Bigg\{\frac{\mathbb{I}(A = 1)}{g(1 \mid W)}
  \frac{q_Z(Z \mid 0, W)}{q_Z(Z \mid 1, W)} -
  \frac{\mathbb{I}(A = 0)}{g(0 \mid W)} \Bigg\} \ .
\end{equation*}
Breaking this down, $\mathbb{I}(A = 1) / g(1 \mid W)$ is the inverse propensity
score weight for $A = 1$ and, likewise, $\mathbb{I}(A = 0) / g(0 \mid W)$ is the
inverse propensity score weight for $A = 0$. The middle term is the ratio of the
conditional densities of the mediator under the control ($A = 0$) and treatment
($A = 1$) conditions (n.b., recall the mediator positivity condition above).

This subtle appearance of a ratio of conditional densities is concerning --
tools to estimate such quantities are sparse in the statistics literature
[@diaz2011super; @hejazi2020haldensify], unfortunately, and the problem is still
more complicated (and computationally taxing) when $Z$ is high-dimensional. As
only the ratio of these conditional densities is required, a convenient
re-parametrization may be achieved, that is,
\begin{equation*}
  \frac{p(A = 0 \mid Z, W)}{g(0 \mid W)}
  \frac{g(1 \mid W)}{p(A = 1 \mid Z, W)} \ .
\end{equation*}
Going forward, we will denote this re-parameterized conditional probability
functional $e(A \mid Z, W) := p(A \mid Z, W)$. The same re-parameterization
technique has been used by @zheng2012targeted, @tchetgen2013inverse,
@diaz2020causal, @diaz2020nonparametric, and @hejazi2021nonparametric in similar
contexts. This reformulation is particularly useful for the fact that it reduces
the estimation problem to one requiring only the estimation of conditional
means, opening the door to the use of a [wide range of machine learning
algorithms, as discussed previously](#sl3).

Underneath the hood, the mean outcome difference $\overline{Q}_{\text{diff}}$
and $e(A \mid Z, W)$, the conditional probability of $A$ given $Z$ and $W$, are
used in constructing the auxiliary covariate for TML estimation. These nuisance
parameters play an important role in the bias-correcting update step of the TML
estimation procedure.

### Estimating the Natural Indirect Effect

Derivation and estimation of the NIE is analogous to that of the NDE. Recall
that the NIE is the effect of $A$ on $Y$ _only through the mediator $Z$_.
This counterfactual quantity, which may be expressed $\E(Y(Z(1), 1) -
\E(Y(Z(0), 1)$, corresponds to the difference of the conditional mean of $Y$
given $A = 1$ and $Z(1)$ (the values the mediator would take under $A = 1$) and
the conditional expectation of $Y$ given $A = 1$ and $Z(0)$ (the values the
mediator would take under $A = 0$).

As with the NDE, re-parameterization can be used to replace $q_Z(Z \mid A, W)$
with $e(A \mid Z, W)$ in the estimation process, avoiding estimation of a
possibly multivariate conditional density.  However, in this case, the mediated
mean outcome difference, previously computed by regressing
$\overline{Q}_{\text{diff}}$ on $W$ among the control units (for whom $A = 0$ is
observed) is instead replaced by a two-step process.  First, $\overline{Q}_Y(Z,
1, W)$, the conditional mean of $Y$ given $Z$ and $W$ when $A = 1$, is regressed
on $W$, among only the treated units (i.e., for whom $A = 1$ is observed). Then,
the same quantity, $\overline{Q}_Y(Z, 1, W)$ is again regressed on $W$, but this
time among only the control units (i.e., for whom $A = 0$ is observed). The mean
difference of these two functionals of the data distribution is a valid
estimator of the NIE.  It can be thought of as the additive marginal effect of
treatment on the conditional mean of $Y$ given $(W, A = 1, Z)$ through its
effect on $Z$. So, in the case of the NIE, while our estimand
$\psi_{\text{NIE}}$ is different, the same estimation techniques useful for
constructing efficient estimators of the NDE come into play.

## The Population Intervention Direct and Indirect Effects

At times, the natural direct and indirect effects may prove too limiting, as
these effect definitions are based on _static interventions_ (i.e., setting
$A = 0$ or $A = 1$), which may be unrealistic for real-world interventions. In
such cases, one may turn instead to the population intervention direct effect
(PIDE) and the population intervention indirect effect (PIIE), which are based
on decomposing the effect of the population intervention effect (PIE) of
flexible stochastic interventions [@diaz2020causal].

We previously discussed stochastic interventions when considering [how to
intervene on continuous-valued treatments](#shift); however, these intervention
schemes may be applied to all manner of treatment variables.
A particular type of stochastic intervention well-suited to working with binary
treatments is the _incremental propensity score intervention_ (IPSI), first
proposed by @kennedy2019nonparametric. Such interventions do not
deterministically set the treatment level of an observed unit to a fixed
quantity (i.e., setting $A = 1$), but instead _alter the odds of receiving the
treatment_ by a fixed amount ($0 \leq \delta \leq \infty$) for each individual.
In particular, this intervention takes the form
\begin{equation*}
  g_{\delta}(1 \mid w) = \frac{\delta g(1 \mid w)}{\delta g(1 \mid w) + 1
  - g(1\mid w)},
\end{equation*}
where the scalar $0 < \delta < \infty$ specifies a _change in the odds of
receiving treatment_. As described by @diaz2020causal in the context of causal
mediation analysis, the identification assumptions required for the PIDE and the
PIIE are significantly more lax than those required for the NDE and NIE. These
identification assumptions include the following. Importantly, the assumption of
cross-world counterfactual independence is not at all required.

::: {.definition name="Conditional Exchangeability of Treatment and Mediators"}
Assume that $\E\{Y(a, z) \mid Z, A, W\} = \E\{Y(a, z) \mid Z,
W\}~\forall~(a, z) \in \mathcal{A} \times \mathcal{Z}W$. This assumption is
stronger than and implied by the assumption $Y(a, z) \indep (A,Z) \mid W$,
originally proposed by @vansteelandt2012natural for identification of mediated
effects among the treated. In introducing this assumption @diaz2020causal state
that "This assumption would be satisfied for any pre-exposure $W$ in a
randomized experiment in which exposure and mediators are randomized. Thus, the
direct effect for a population intervention corresponds to contrasts between
treatment regimes of a randomized experiment via interventions on $A$ and $Z$,
unlike the natural direct effect for the average treatment effect
[@robins2010alternative]."
:::

::: {.definition name="Common Support of Treatment and Mediators"}
Assume that $\text{supp}\{g_{\delta}(\cdot \mid w)\} \subseteq
\text{supp}\{g(\cdot \mid w)\}~\forall~w \in \mathcal{W}$. This assumption is
standard and requires only that the post-intervention value of $A$ be supported
in the data. Note that this is significantly weaker than the treatment and
mediator positivity conditions required for the natural direct and indirect
effects, and it is a direct consequence of using stochastic (rather than static)
interventions.
:::

## Evaluating the Direct and Indirect Effects

We now turn to estimating the natural direct and indirect effects, as well as
the population intervention direct effect, using the WASH Benefits data,
introduced in earlier chapters. Let's first load the data:

```{r tmle3mediate-load-data}
library(data.table)
library(sl3)
library(tmle3)
library(tmle3mediate)

# download data
washb_data <- fread(
  paste0(
    "https://raw.githubusercontent.com/tlverse/tlverse-data/master/",
    "wash-benefits/washb_data.csv"
  ),
  stringsAsFactors = TRUE
)

# make intervention node binary and subsample
washb_data <- washb_data[sample(.N, 600), ]
washb_data[, tr := as.numeric(tr != "Control")]
```

We'll next define the baseline covariates $W$, treatment $A$, mediators $Z$,
and outcome $Y$ nodes of the NPSEM via a "Node List" object:

```{r tmle3mediate-node-list}
node_list <- list(
  W = c(
    "momage", "momedu", "momheight", "hfiacat", "Nlt18", "Ncomp", "watmin",
    "elec", "floor", "walls", "roof"
  ),
  A = "tr",
  Z = c("sex", "month", "aged"),
  Y = "whz"
)
```

Here, the `node_list` encodes the parents of each node -- for example, $Z$ (the
mediators) have parents $A$ (the treatment) and $W$ (the baseline confounders),
and $Y$ (the outcome) has parents $Z$, $A$, and $W$. We'll also handle any
missingness in the data by invoking `process_missing`:

```{r tmle3mediate-process_missing}
processed <- process_missing(washb_data, node_list)
washb_data <- processed$data
node_list <- processed$node_list
```

We'll now construct an ensemble learner using a handful of popular machine
learning algorithms:

```{r tmle3mediate-sl-learners}
# SL learners used for continuous data (the nuisance parameter Z)
enet_contin_learner <- Lrnr_glmnet$new(
  alpha = 0.5, family = "gaussian", nfolds = 3
)
lasso_contin_learner <- Lrnr_glmnet$new(
  alpha = 1, family = "gaussian", nfolds = 3
)
fglm_contin_learner <- Lrnr_glm_fast$new(family = gaussian())
mean_learner <- Lrnr_mean$new()
contin_learner_lib <- Stack$new(
  enet_contin_learner, lasso_contin_learner, fglm_contin_learner, mean_learner
)
sl_contin_learner <- Lrnr_sl$new(learners = contin_learner_lib)

# SL learners used for binary data (nuisance parameters G and E in this case)
enet_binary_learner <- Lrnr_glmnet$new(
  alpha = 0.5, family = "binomial", nfolds = 3
)
lasso_binary_learner <- Lrnr_glmnet$new(
  alpha = 1, family = "binomial", nfolds = 3
)
fglm_binary_learner <- Lrnr_glm_fast$new(family = binomial())
binary_learner_lib <- Stack$new(
  enet_binary_learner, lasso_binary_learner, fglm_binary_learner, mean_learner
)
sl_binary_learner <- Lrnr_sl$new(learners = binary_learner_lib)

# create list for treatment and outcome mechanism regressions
learner_list <- list(
  Y = sl_contin_learner,
  A = sl_binary_learner
)
```

### Targeted Estimation of the Natural Indirect Effect

We demonstrate calculation of the NIE below, starting by instantiating a "Spec"
object that encodes exactly which learners to use for the nuisance parameters
$e(A \mid Z, W)$ and $\psi_Z$. We then pass our Spec object to the `tmle3`
function, alongside the data, the node list (created above), and a learner list
indicating which machine learning algorithms to use for estimating the nuisance
parameters based on $A$ and $Y$.

```{r tmle3mediate-NIE, eval = FALSE}
tmle_spec_NIE <- tmle_NIE(
  e_learners = Lrnr_cv$new(lasso_binary_learner, full_fit = TRUE),
  psi_Z_learners = Lrnr_cv$new(lasso_contin_learner, full_fit = TRUE),
  max_iter = 1
)
washb_NIE <- tmle3(
  tmle_spec_NIE, washb_data, node_list, learner_list
)
washb_NIE
```

Based on the output, we conclude that the indirect effect of the treatment
through the mediators `node_list$Z` is `washb_NIE$summary$tmle_est`.

### Targeted Estimation of the Natural Direct Effect

An analogous procedure applies for estimation of the NDE, only replacing the
Spec object for the NIE with `tmle_spec_NDE` to define learners for the NDE
nuisance parameters:

```{r tmle3mediate-NDE, eval = FALSE}
tmle_spec_NDE <- tmle_NDE(
  e_learners = Lrnr_cv$new(lasso_binary_learner, full_fit = TRUE),
  psi_Z_learners = Lrnr_cv$new(lasso_contin_learner, full_fit = TRUE),
  max_iter = 1
)
washb_NDE <- tmle3(
  tmle_spec_NDE, washb_data, node_list, learner_list
)
washb_NDE
```

From this, we can draw the conclusion that the direct effect of the treatment
(through all paths not involving the mediators, `node_list$Z`) is
`washb_NDE$summary$tmle_est`. Note that, together, the estimates of
the natural direct and indirect effects approximately recover the _average
treatment effect_, that is, based on these estimates of the NDE and NIE, the
ATE is roughly the sum of `washb_NDE$summary$tmle_est` and 
`washb_NIE$summary$tmle_est`.

## Exercises

::: {.exercise}
Examine the WASH Benefits dataset and choose a different set of potential
mediators of the effect of the treatment on weight-for-height Z-score. Using
this newly chosen set of mediators (or single mediator), estimate the natural
direct and indirect effects. Provide an interpretation of these estimates.
:::

::: {.exercise}
Assess whether additivity of the natural direct and indirect effects holds.
Using the natural direct and indirect effects estimated above, does their sum
recover the ATE?
:::

::: {.exercise}
Evaluate whether the assumptions required for identification of the natural
direct and indirect effects are plausible in the WASH Benefits example. In
particular, position this evaluation in terms of empirical diagnostics of
treatment and mediator positivity.
:::
